---
title: "Differential Expression Metabolites"
output: html_document
---

# Analysis to detect metabolic change after sleep deprivation

We aim to detect the metabolites that significantly change after sleep deprivation. Our dataset contain the metabolic concentration value for BXD mice, parental and F1.

As our BXD lines respond differently to sleep deprivation, we include in our model the specific lines effect.


We read the file

```{r}
File<-"../Data/Metabolite_BXD.txt"
x<-read.table(File,header=T,sep="\t")
n<-x[1,] # Contains the metabolites type
names(n)<-gsub(".","_",names(n),fixed = TRUE) # reformate a bit the data
x<-x[-1,] # We remove the first lines containing the metabolites type
```

Some of our results contains Null values (-1), we replace them by NA values

```{r}
head(x[1:10,1:10])
x[x==-1]<-NA
```

We compute differential expression pvalue using 2 factor anova:

We use the lines effect (Group) and the Condition effect

We compute also the global fold change effect of sleep deprivation.

```{r}
cpval<-c()
fc<-c()
for (i in 6:ncol(x)){ # we only start at column 6. (previous column contains other information)
  cpval<-c(cpval,summary(aov(as.numeric(as.character(x[,i]))~factor(x$Group)+factor(x$Condition)))[[1]][2,"Pr(>F)"])
  fit<-lm(as.numeric(as.character(x[,i]))~factor(x$Group)+factor(x$Condition)) 
  coefsd<-fit$coefficients["factor(x$Condition)SD"]
  intercept<-fit$coefficients["(Intercept)"]
  fc<-c(fc,(intercept+coefsd)/intercept)
}
names(cpval)<-colnames(x)[6:ncol(x)]
names(fc)<-colnames(x)[6:ncol(x)]
```

We correct for false discovery rate

```{r}
cpval<-p.adjust(cpval,method="fdr")
```


We rename a few metabolites for the figures

```{r}
n6<-names(n)[seq(41,48)]
n7<-names(n)[seq(49,79)]
n8<-names(n)[seq(80,115)]
n[n6]="Lysophosphatidylcholine"
n[n7]="Phosphatidylcholine (diacyl)"
n[n8]="Phosphatidylcholine (acyl-alkyl)"
nn<-c()
for (i in 6:length(n)){
	nn<-c(nn,as.character(n[1,i]))
}
```

We save our results

```{r}
MM<-matrix(nrow=124,ncol=4)
MM[,1]<-names(cpval)
MM[,2]<-nn
MM[,3]<-cpval
MM[,4]<-fc

colnames(MM)<-c("Metabolite","Type","adjPval","FC")
MM<-as.data.frame(MM)
write.table(MM,file="Differential_Expression_Metabolites_Results.txt",quote=F,sep="\t")
```

We plot our results (Figure from the paper)

```{r}
library(ggplot2)
g<-ggplot(MM,aes(-log10(as.numeric(as.character(MM$adjPval))),log2(as.numeric(as.character(MM$FC)))))+geom_point(aes(colour=MM$Type),size=4)+scale_colour_brewer(name="Metabolites",palette = "Set1")+
  geom_vline(xintercept=-log10(0.05),col="red",size=2)+scale_x_reverse()+xlab("log FDR adjusted p-value")+ylab("log fold−change")+
  theme(panel.background = element_rect(fill = rgb(0,0,0,alpha=0.1), colour = 'black'))+
  geom_hline(aes(yintercept=0),col="black",size=1)
g
pdf("Metabolite_DE.pdf",height = 8, width = 8)
g
dev.off()
```

```{r}
nrow(MM[as.numeric(as.character(MM$adjPval))<=0.05,])
```

```{r}
MM[as.numeric(as.character(MM$adjPval))<=0.05,]
```



