---
title: "Heritability"
output:
  pdf_document:
    toc: no
  html_document:
    code_folding: hide
    df_print: paged
    rows.print: 6
    theme: sandstone
    toc: yes
    toc_float: yes
---

```{r,echo=T,warning=F,error=F,message=F}
PATH<-"../../"
setwd(paste(PATH,"Analysis/BXD_Paper_Figures",sep=""))
library(ggplot2)
```

# Heritability informations:

* The percent variance accounted for by strain is considered by some to be an estimate of broad-sense heritability (Hirsch 1967; Lynch & Walsh 1998) for clones, and is formulated as the strain intra-class correlation: [http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2855868/]

* Broad-sense (H2) and narrow-sense (h2) H2 = Va/Vt; h2 = 1/2Va/(1/2Va + Vw) was estimated using the Hegmann and Possidente Method (1981)http://journal.frontiersin.org/article/10.3389/fnbeh.2014.00341/full, http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4204525/ 

# Heritability function

The function return a matrix with: 

* FileName

* Phenotypes Name 

* Narrow-sense heritability 

* Broad-sense heritability

* FileName + Phenotype Name

```{r}
# Heritability for Sleep phenotypes
Heritability<-function(File){
  m<-matrix(ncol=5,nrow=0)
  
  # Remove parental lines and F1, take only BXD mice
  x<-read.table(File,header=T)
  
  #x<-x[-which(as.character(x$line) %in% c("C57BL6","DBA2","BXDF1","DXBF1")),]
  x<-x[-which(as.character(x$line) %in% c("BXD005","BXD029","BXD029TL","BXD032","BXD063","BXD029TLR4")),]

  #if (length(unique(x$line))>33){print(unique(x$line))}
  
  # change file name
  File<-strsplit(File,"/")
  File<-File[[1]][length(File[[1]])]
  
  for (i in seq(8,ncol(x))){
    a<-aov(as.numeric(as.character(x[,i])) ~ x$line)
    Va<-summary(a)[[1]][1,3]  #is the variance between strains (Mean square for lines effect)
    Vw<-summary(a)[[1]][2,3] #is the variance within strains. (Mean square for residuals)
    Vt<-Va+Vw # Total mean variance
    
    # Broad-sense heritability
    H2<-Va/Vt
    # Narrow-sense heritability
    h2<-0.5*Va/(0.5*Va+Vw)
    
    m<-rbind(m,c(File,colnames(x)[i],h2,H2,paste(File,colnames(x)[i],sep='.')))
  }
  return(m)
}

# Heritability for metabolites
HeritabilityM<-function(File){
  m<-matrix(ncol=5,nrow=0)
  x<-read.table(File,header=T,sep="\t",stringsAsFactors = F)
  
  # NSD
  ## Remove NA
  x[x==(-1)]<-NA
  
  ## filter data
  n<-x[1,]
  x<-x[-1,]
  
  #x<-x[-which(as.character(x$Group) %in% c("C57BL6","DBA","D2B6F1","B6D2F1")),]
  x<-x[-which(as.character(x$Group) %in% c("BXD05","BXD29","BXD29TLR4","BXD32","BXD63")),]

  
  #print(length(unique(x$Group)))
  
  xNSD<-x[x$Condition=="NSD",]
  
  for (i in seq(6,ncol(xNSD))){
    a<-aov(as.numeric(as.character(xNSD[,i])) ~ xNSD$Group)

    Va<-summary(a)[[1]][1,3] #is the variance between strains (Mean square for lines effect)
    Vw<-summary(a)[[1]][2,3] #is the variance within strains. (Mean square for residuals)
    Vt<-Va+Vw
    
    H2<-Va/Vt
    h2<-0.5*Va/(0.5*Va+Vw)

    m<-rbind(m,c("NSD",colnames(xNSD)[i],h2,H2,as.character(n[i][[1]])))
  }
  
  # SD
  xSD<-x[x$Condition=="SD",]
  for (i in seq(6,ncol(xSD))){
    a<-aov(as.numeric(as.character(xSD[,i])) ~ xSD$Group)
    
    Va<-summary(a)[[1]][1,3] #is the variance between strains (Mean square for lines effect)
    Vw<-summary(a)[[1]][2,3] #is the variance within strains. (Mean square for residuals)
    Vt<-Va+Vw
    
    H2<-Va/Vt
    h2<-0.5*Va/(0.5*Va+Vw)

    m<-rbind(m,c("SD",colnames(xSD)[i],h2,H2,as.character(n[i][[1]])))
  }
  
  # Rec vs Bsl
  for (i in seq(6,ncol(xSD))){
    
    # Mean NSD
    xmNSDtmp<-aggregate(as.numeric(as.character(xNSD[,i])),list(xNSD$Group),mean,na.rm=T)
    xmNSD<-xmNSDtmp$x
    names(xmNSD)<-xmNSDtmp$Group.1
    
    xSDtmp<-log2(as.numeric(as.character(xSD[,i]))/xmNSD[as.character(xSD$Group)])
    #xSDtmp<-as.numeric(as.character(xSD[,i]))-xmNSD[as.character(xSD$Group)]
    #xSDtmp<-as.numeric(as.character(xSD[,i]))/xmNSD[as.character(xSD$Group)]
    a<-aov(xSDtmp ~ xSD$Group)
    
    Va<-summary(a)[[1]][1,3] #is the variance between strains (Mean square for lines effect)
    Vw<-summary(a)[[1]][2,3] #is the variance within strains. (Mean square for residuals)
    Vt<-Va+Vw
    
    H2<-Va/Vt
    h2<-0.5*Va/(0.5*Va+Vw)
    
    m<-rbind(m,c("RECvsBSL",colnames(x)[i],h2,H2,as.character(n[i][[1]])))
  }
  
  return(m)
}
```


# Computation

Get Heritability for EEG/Behavioral phentoypes

```{r}
# Heritability Matrix
aH <- matrix(nrow=0,ncol=5)

# Compute heritability for all phenotypes files
for (File in list.files(paste(PATH,"Rdata/Phenotypes/",sep=""),pattern='*quant.out*')){
  aH<-rbind(aH,Heritability(paste(PATH,"Rdata/Phenotypes/",File,sep="")))
}

# Read phenotype Categories files & rename a few phenotypes
categories<-read.table(paste(PATH,"Rdata/Phenotype.Categories.txt",sep=""),header=T,sep="\t")
categories$filename.variablename<-gsub("BXDvs1224bsl.","BXDvs1224bsl.out.",categories$filename.variablename)
categories$filename.variablename<-gsub(".out.",".quant.out.",categories$filename.variablename)
categories$filename.variablename<-gsub("BXD_TPF.P","BXD_TPF-P",categories$filename.variablename)
categories$filename.variablename<-gsub("n6hloss","n6hLoss",categories$filename.variablename)
categories$filename.variablename<-gsub("n6hloss","n6hLoss",categories$filename.variablename)
categories$filename.variablename<-gsub("n24hgain","n24hGain",categories$filename.variablename)
categories$filename.variablename<-gsub("r6hloss","r6hLoss",categories$filename.variablename)
categories$filename.variablename<-gsub("r24hgain","r24hGain",categories$filename.variablename)

# Transform into dataframe
h2df<-as.data.frame(aH)
h2df$V3<-as.numeric(as.character(h2df$V3))

# Add categories
h2df<-h2df[h2df$V5 %in% categories$filename.variablename,]
rownames(categories)<-categories$filename.variablename
m<-matrix(nrow=0,ncol=3)
for (i in h2df$V5){
  m<-rbind(m,(categories[categories$filename.variablename == i,c(2,3,4)]))
}
h2df<-cbind(h2df,m)
colnames(h2df)<-c("File","Phenotype","h2","H2","FilePhenotype","Condition","Categories","Subcategories")
```

Get Heritability for metabolites

```{r}
# Get heritability
mH<-HeritabilityM(paste(PATH,"Rdata/Metabolites_BXD_alldata.txt",sep=""))
mH<-as.data.frame(mH)
mH2<-mH[mH$V1=="NSD",]

h2df<-as.matrix(h2df)

# Metabolite NSD
for (i in 1:nrow(mH2)){
  h2df<-rbind(h2df,c(as.character(mH2[i,2]),
                 as.character(mH2[i,2]),
                 as.character(mH2[i,3]),
                 as.character(mH2[i,4]),
                 as.character(mH2[i,2]),
                 "bsl",
                 "ZMetabolite",
                 as.character(mH2[i,5])
                 ))
}

# Metabolite SD
mH2<-mH[mH$V1=="SD",]
for (i in 1:nrow(mH2)){
  h2df<-rbind(h2df,c(as.character(mH2[i,2]),
                 as.character(mH2[i,2]),
                 as.character(mH2[i,3]),
                 as.character(mH2[i,4]),
                 as.character(mH2[i,2]),
                 "rec",
                 "ZMetabolite",
                 as.character(mH2[i,5])
                 ))
}

# Metabolite rec vs bsl
mH2<-mH[mH$V1=="RECvsBSL",]
for (i in 1:nrow(mH2)){
  h2df<-rbind(h2df,c(as.character(mH2[i,2]),
                 as.character(mH2[i,2]),
                 as.character(mH2[i,3]),
                 as.character(mH2[i,4]),
                 as.character(mH2[i,2]),
                 "rec vs bsl",
                 "ZMetabolite",
                 as.character(mH2[i,5])
                 ))
}


h2df<-as.data.frame(h2df)
h2df$h2<-as.numeric(as.character(h2df$h2))
head(h2df)
```


```{r}
write.table(file="h2values.txt",h2df,sep="\t",col.names = T,row.names = T,quote = F)
```


# Plot results

```{r,fig.height=10,fig.width=15}
h2df$UniqueCatSubCat<-paste(h2df$Categories,h2df$Subcategories,sep="_")

theme_set(theme_gray(base_size = 25))
ggplot(data=h2df,aes(UniqueCatSubCat,h2))+geom_violin(aes(fill=Categories))+geom_jitter(aes(color=Condition),size=3.5,alpha=0.5,height = NULL,width=.15)+
xlab("")+ylab("heritability")+ylim(0.2,1)+
scale_colour_brewer(palette="Set1",breaks=c("bsl","rec","rec vs bsl","SD"),
                    labels=c("Baseline","Recovery","Recovery vs Baseline","Sleep Deprivation"),
                    name="Condition")+
scale_fill_brewer(palette = "Pastel1",name="Type",breaks=c("activity","EEG","state","ZMetabolite"),labels=c("Activity","EEG","State","Metabolite"))+
theme(axis.text.x = element_text(angle = 45, hjust = 1))+
scale_x_discrete(labels=c("Amount","Distribution","Frequency",
                          "Paroxismal","Power","Amount","Architecture",
                          "Distribution","ACT","AA",
                          "BA","GPL","SM"))
```


```{r}
pdf(file="Heritability.pdf",height = 10,width = 15)
theme_set(theme_gray(base_size = 25))
ggplot(data=h2df,aes(UniqueCatSubCat,h2))+geom_violin(aes(fill=Categories))+geom_jitter(aes(color=Condition),size=3.5,alpha=0.5,height = NULL,width=.15)+
xlab("")+ylab("heritability")+ylim(0.2,1)+
scale_colour_brewer(palette="Set1",breaks=c("bsl","rec","rec vs bsl","SD"),
                    labels=c("Baseline","Recovery","Recovery vs Baseline","Sleep Deprivation"),
                    name="Condition")+
scale_fill_brewer(palette = "Pastel1",name="Type",breaks=c("activity","EEG","state","ZMetabolite"),labels=c("Activity","EEG","State","Metabolite"))+
theme(axis.text.x = element_text(angle = 45, hjust = 1))+
scale_x_discrete(labels=c("Amount","Distribution","Frequency",
                          "Paroxismal","Power","Amount","Architecture",
                          "Distribution","ACT","AA",
                          "BA","GPL","SM"))
dev.off()
```

# Session information for this run

```{r}
sessionInfo()
```
